---
title: "Tutorial: Central Posterior Envelopes (CPEs) for Functional Components in Bayesian Longitudinal Functional Principal Component (B-LFPCA) Model"
output: html_document
---

## Description 
A step-by-step implementation of the functional depth based central posterior envelopes (CPEs) for posterior functional components in the Bayesian longitudinal functional principal component analysis (B-LFPCA) model detailed in "Central posterior envelopes for Bayesian longitudinal functional principal component analysis" by Boland et al. (2024). 
This procedure assumes that the data is comprised of densely observed longitudinal functional observations. 

## Functions Implemented
B_LFPCA_CPE_simulateData.R, B_LFPCA_CPE_MCMC.R, B_LFPCA_CPE_postSumms.R, B_LFPCA_CPEcontour.R, and B_LFPCA_muCPEcontour.R

## Tutorial Outline 
1. Simulate longitudinal functional data using the product FPCA model (B_LFPCA_CPE_simulateData.R)
2. MCMC to obtain posterior estimates (B_LFPCA_CPE_MCMC.R)
3. Calculate traditional and proposed depth-based posterior summaries (B_LFPCA_CPE_postSumms.R)
4. Visualization of CPE contours for functional model components, including the mean function, longitudinal/functional eigenfunctions (B_LFPCA_CPE_muCPEcontour.R, B_LFPCA_CPE_CPEcontour.R)

```{r, warning = FALSE, message = FALSE}
# Install missing packages
list.of.packages <- c('gridExtra', 'cowplot', 'latex2exp', 'tidyverse', 'reshape2',
                      'viridis', 'splines', 'Rcpp', 'acid', 'pracma', 'expm', 'MASS',
                      'LFBayes')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages) 

# Load packages
library(gridExtra)
library(cowplot)
library(latex2exp)
library(tidyverse)
library(reshape2)
library(viridis)
library(splines)
library(Rcpp)
library(acid)
library(pracma)
library(expm)
library(MASS)
library(LFBayes)

# Set work directory
setwd("~/Desktop/Github")

# Load R functions
source("B_LFPCA_CPE_simulateData.R")
source("B_LFPCA_CPE_MCMC.R")
source("B_LFPCA_CPE_postSumms.R")
source("B_LFPCA_CPE_CPEcontour.R")
source("B_LFPCA_CPE_muCPEcontour.R")
```

## 1. Simulate functional data using the standard FPCA model

In this tutorial, the generated function data contain observations for $n = 30$ 
subjects comprised of values $Y_i(s_j,t_k)$ at $SS = 20$ equidistant longitudinal
time points within $s \in [0, 1]$ and $TT = 20$ equidistant functinal time points
within$t \in [0, 1]$. The longitudinal functional data are generated from the 
standard product FPCA model corresponding  to the simulation case 1 in which no 
outliers are generated. The model components used to generate the data are detailed 
in Supplementary Materials Appendix C. 

```{r, warning = FALSE, message = FALSE}
# Define parameters in the simulated model in Case 1
# Number of longitudinal and functional time points
SS <- 20; TT <- 20

# Longitudinal and functional time points
s <- seq(from = 0, to = 1, length.out = SS) # vector, SS x 1
t <- seq(from = 0, to = 1, length.out = TT) # vector, TT x 1

# Number of longitudinal and functional eigen-components
J <- 2; K <- 2

# Number of subjects 
n <- 30

set.seed(999)
simulatedData <- simulateData(n  = 30, # Number of subjects (scalar)
                              SS = 20, # Number of functional time points (scalar)
                              TT = 20) # Number of longitudinal time points (scalar)
# Simulated data
Y <- simulatedData$y

# True functions
truth_funcs <- simulatedData$truth_funcs
```

## 2. MCMC to obtain posterior estimates 

We obtain the $C = 6000$ posterior estimates for the model parameters and components 
utilizing the B-LFPCA modeling framework using $p_1 = p_2 = 11$ B-spline basis functions and 
$q_1 = q_2 = 4$ latent components. The posterior estimates are obtained from the Gibbs sampler 
described in Supplementary Materials Appendix A using 4 MCMC chains, with each consisting of 
10,000 MCMC iterations. Of these, the first 2,500 iterations were discarded as burn-in, and 
thinning was applied every 5th iteration. 

```{r, cache = TRUE, warning = FALSE, message = FALSE}
# Note: For n = 30, SS = 20, TT = 20, p_1 = p_2 = 11, and q_1 = q_2 = 4, running four MCMC 
#       chain with 10,000 iterations takes approximately 1 hour

# MCMC parameters
## Number of B-spline basis functions
p1 <- p2 <- 11

## Number of latent components
q1 <- q2 <- 4

## Number of MCMC chains, number of total iterations, iterations used as burn-in, and thinning number in each chain
nchain <- 4
iter <- 10000
burnin <- 2500
thin <- 5

# MCMC
posterior_samples <-   MCMC(Y,           # List of simulated functional data, each list contains a vector of Y_i(s, t)                                              (list, (SS x TT) x 1 x n )
                            n,           # Number of subjects (scalar)
                            SS,          # Number of functional time points (scalar)
                            TT,          # Number of longitudinal time points (scalar)
                            p1,          # Number of longitudinal basis functions (integer)
                            p2,          # Number of functional basis functions (integer)
                            q1,          # Number of longitudinal latent functions (integer)
                            q2,          # Number of functional latent functions (integer)
                            nchain,      # Number of MCMC chains (integer)
                            iter,        # Total number of iterations in each MCMC chain (integer)
                            burnin,      # Number of iterations used for burn-in (integer, 0-iter)
                            thin)        # Number of thinning (integer)

```

## 3. Calculate traditional and proposed posterior summaries 

```{r, cache = TRUE, warning = FALSE, message = FALSE}
# Note: For n = 30, SS = 20, TT = 20, p_1 = p_2 = 11, and q_1 = q_2 = 4, running summaries with 6,000 posterior samples takes approximately 4 hours

# Set parameters
# Number of eigencomponent retained
neig <- 2

# Alpha-level
alpha <- 0.05

# Calculate posterior summaries
posterior_summaries <- postSumms(posterior_samples, # Output object from MCMC() of FDpostSumms_BLFPCA_MCMC.R
                                 neig,              # Number of eigencomponents (integer) 
                                 n,                 # Number of subjects (scalar)
                                 SS,                # Number of functional time points (scalar)
                                 TT,                # Number of longitudinal time points (scalar)
                                 p1,                # Number of longitudinal basis functions (integer)
                                 p2,                # Number of functional basis functions (integer)
                                 q1,                # Number of longitudinal latent functions (integer)
                                 q2,                # Number of functional latent functions (integer)
                                 nchain,            # Number of MCMC chains (integer)
                                 iter,              # Total number of iterations in each MCMC chain (integer)
                                 burnin,            # Number of iterations used for burn-in (integer, 0-iter)
                                 thin,              # Number of thinning (integer)
                                 alpha              # Alpha-level of (1 - alpha)100% credible intervals and central                                                            posterior envelopes (scalar, 0-1)
)                           
```

## 4. CPE contour visualizations for functional model components

The following codes are make CPE contour plots for different types of CPE contours for eigenfunctions in Case 1 for a single simulation ran. In the figures, the posterior sample is displayed as solid grey lines, and each $alpha$-level contour corresponds for the corresponding MBD- or marginal MVD- or kernel MVD-CPEs to the color given in the legend. The solid black line represents either the MBD median or MVD median. 

### longitudinal Eigenfunctions

```{r, warning = FALSE, message = FALSE}
contour_case1_psi1_mbd <- CPEcontour(
                       post_samples = posterior_summaries,      # Outputs from FDpostSumms_BLFPCA_postSumms.R
                       type = "long_eigen",                     # Type of eigenfunctions: "long_eigen" or "func_eigen".
                       jork = 1,                                # Index for the eigenfunctions: 1, 2, or 3
                       CPE = "MBD",                             # Type of CPE calculated and plotted: "MBD", "Marginal                                                                     MVD", or "Kernel MVD"    
                       alpha_cont = seq(0.05, 0.95, by = 0.10), # alpha values for the CPE contours, a vector of real                                                                      numbers
                       sort = s,                                # To denote the longitudinal (s) or functional (t)                                                                         dimension
                       title = TeX(r"((a) Case 1, MBD-CPEs)"),  # Title given to the plot, charactors
                       xlab = "Time (s)"                        # X-axis label, characters
                       )

contour_case1_psi1_mvd <- CPEcontour(post_samples = posterior_summaries, 
                                     type = "long_eigen", 
                                     jork = 1, 
                                     CPE = "Marginal MVD", 
                                     alpha_cont = seq(0.05, 0.95, by = 0.10), 
                                     sort = s, 
                                     title = TeX(r"((b) Case 1, MVD-CPEs)"), 
                                     xlab = "Time (s)")

contour_case1_psi1_kmvd <- CPEcontour(post_samples = posterior_summaries, 
                                      type = "long_eigen", 
                                      jork = 1, 
                                      CPE = "Kernel MVD", 
                                      alpha_cont = seq(0.05, 0.95, by = 0.10), 
                                      sort = s, 
                                      title = TeX(r"((c) Case 1, Kernel MVD-CPEs)"), 
                                      xlab = "Time (s)")

contour_case1_psi2_mbd <- CPEcontour(post_samples = posterior_summaries, 
                                     type = "long_eigen", 
                                     jork = 2, 
                                     CPE = "MBD", 
                                     alpha_cont = seq(0.05, 0.95, by = 0.10), 
                                     sort = s, 
                                     title = TeX(r"((a) Case 1, MBD-CPEs)"), 
                                     xlab = "Time (s)")

contour_case1_psi2_mvd <- CPEcontour(post_samples = posterior_summaries, 
                                     type = "long_eigen", 
                                     jork = 2, 
                                     CPE = "Marginal MVD", 
                                     alpha_cont = seq(0.05, 0.95, by = 0.10), 
                                     sort = s, 
                                     title = TeX(r"((b) Case 1, MVD-CPEs)"), 
                                     xlab = "Time (s)")

contour_case1_psi2_kmvd <- CPEcontour(post_samples = posterior_summaries, 
                                      type = "long_eigen", 
                                      jork = 2, 
                                      CPE = "Kernel MVD", 
                                      alpha_cont = seq(0.05, 0.95, by = 0.10), 
                                      sort = s, 
                                      title = TeX(r"((c) Case 1, Kernel MVD-CPEs)"), 
                                      xlab = "Time (s)")

```


### Functional Eigenfunctions

```{r, warning = FALSE, message = FALSE}
contour_case1_phi1_mbd <- CPEcontour(post_samples = posterior_summaries, 
                                     type = "func_eigen", 
                                     jork = 1, 
                                     CPE = "MBD", 
                                     alpha_cont = seq(0.05, 0.95, by = 0.10),
                                     sort = t, 
                                     title = TeX(r"((a) Case 1, MBD-CPEs)"), 
                                     xlab = "Time (t)")

contour_case1_phi1_mvd <- CPEcontour(post_samples = posterior_summaries, 
                                     type = "func_eigen", 
                                     jork = 1, 
                                     CPE = "Marginal MVD", 
                                     alpha_cont = seq(0.05, 0.95, by = 0.10),
                                     sort = t, 
                                     title = TeX(r"((b) Case 1, MVD-CPEs)"), 
                                     xlab = "Time (t)")

contour_case1_phi1_kmvd <- CPEcontour(post_samples = posterior_summaries, 
                                      type = "func_eigen", 
                                      jork = 1, 
                                      CPE = "Kernel MVD", 
                                      alpha_cont = seq(0.05, 0.95, by = 0.10), 
                                      sort = t, 
                                      title = TeX(r"((c) Case 1, Kernel MVD-CPEs)"), 
                                      xlab = "Time (t)")

contour_case1_phi2_mbd <- CPEcontour(post_samples = posterior_summaries, 
                                     type = "func_eigen", 
                                     jork = 2, 
                                     CPE = "MBD", 
                                     alpha_cont = seq(0.05, 0.95, by = 0.10), 
                                     sort = t, 
                                     title = TeX(r"((a) Case 1, MBD-CPEs)"), 
                                     xlab = "Time (t)")

contour_case1_phi2_mvd <- CPEcontour(post_samples = posterior_summaries, 
                                     type = "func_eigen", 
                                     jork = 2, 
                                     CPE = "Marginal MVD", 
                                     alpha_cont = seq(0.05, 0.95, by = 0.10), 
                                     sort = t, 
                                     title = TeX(r"((b) Case 1, MVD-CPEs)"), 
                                     xlab = "Time (t)")

contour_case1_phi2_kmvd <- CPEcontour(post_samples = posterior_summaries, 
                                      type = "func_eigen", 
                                      jork = 2, 
                                      CPE = "Kernel MVD", 
                                      alpha_cont = seq(0.05, 0.95, by = 0.10), 
                                      sort = t, 
                                      title = TeX(r"((c) Case 1, Kernel MVD-CPEs)"), 
                                      xlab = "Time (t)")

```


### Visualization of MVD-CPE contours of the mean function 

Plotted at different $\alpha$ levels marked by varying colors, overlaid with $C = 6000$ posterior estimates in gray. To visualize the MVD-CPE contours, the CPEs of the mean function are displayed in functional time (t) at fixed
slices of longitudinal time $s \in \mathcal{S}$ (i.e. ($\mu(t|s)$) for the $n_s = 20$ total time points. The solid black line represents the MVD median, denoted by $\widetilde{m}\{\mu(s, t)\}$, and the solid red line represents the true function.

```{r, warning = FALSE, message = FALSE, fig.height=15, fig.width=18}
muCPEcontour(post_samples = posterior_summaries,        # Outputs from FDpostSumms_BLFPCA_postSumms.R
               alpha_cont = seq(0.05, 0.95, by = 0.10), # Vector of alpha values for the CPE contours, a vector of real                                                            numbers
                        s = seq(0, 1, length = 20),     # Vector of longitudinal time points to be plotted, a                                                                      vector of real numbers
                        t = seq(0, 1, length = 20)      # Vector of functional time points to be plotted, a                                                                        vector of real numbers
)
```


### Visualization of CPE contours of longtudinal and functional eigenfunctions

```{r, warning = FALSE, message = FALSE, fig.height=11, fig.width=12}
cols <- viridis::viridis(n = 10)
# Generate a dataset and plot to obtain legend
df <- data.frame(x = rnorm(1000, sd = 100), y = rnorm(1000, sd = 100))
grid <- ggplot(df, aes(x = x, y = y)) +
  geom_density_2d_filled(bins = 10) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title=element_text(size=12),
        legend.text = element_text(size = 11),
        legend.title = element_text(size = 13),
        legend.background = element_rect(colour = 'black', fill = 'white', 
                                         linetype='solid'),
        axis.text=element_text(size=12)) +
  scale_fill_manual(values =  cols, 
                    labels = c(as.character(seq(0.05, 0.95, by = 0.1)))) +
  guides(fill=guide_legend(title=TeX(r'($\alpha$ level)')))
legend <- get_legend(grid)

# Make plots of CPE contours
ggdraw() +
  draw_plot(contour_case1_psi1_mbd , x = 0, y = 1 - 1/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_psi1_mvd , x = 0.3, y = 1 - 1/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_psi1_kmvd , x = 0.6, y = 1 - 1/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_psi2_mbd , x = 0, y = 1 - 2/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_psi2_mvd , x = 0.3, y = 1 - 2/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_psi2_kmvd , x = 0.6, y = 1 - 2/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_phi1_mbd , x = 0, y = 1 - 3/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_phi1_mvd , x = 0.3, y = 1 - 3/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_phi1_kmvd , x = 0.6, y = 1 - 3/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_phi2_mbd , x = 0, y = 1 - 4/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_phi2_mvd , x = 0.3, y = 1 - 4/4, width = 0.3 - 0.01, height = 1/4) +
  draw_plot(contour_case1_phi2_kmvd , x = 0.6, y = 1 - 4/4, width = 0.3 - 0.01, height = 1/4) 
  draw_grob(legend, x = 0.9, y = 0.25, width = 0.1 - 0.01, height = 0.5)

```

